### Raster structure

Questions:
- What is a raster dataset?
- How do I work with and plot raster data in R?
- How can I handle missing or bad data values for a raster?

Objectives:
- Describe the fundamental attributes of a raster dataset.
- Explore raster attributes and metadata using R.
- Import rasters into R using the raster package.
- Plot a raster file in R using the ggplot2 package.
- Describe the difference between single- and multi-band rasters.


Installing the raster and rgdal packages: install.packages("raster").

Note that for raster package you might need to install sp package using install.packages("sp") as well as install.packages("rgdal")

```{r loadLibraries}
library(raster)
library(rgdal)
library(tidyverse)
# set the ggplot theme to minimal upfront
theme_set(theme_minimal())
```

You should already have the data...

First, let's view the Raster file attributes

We will be working with a series of GeoTIFF files in this lesson. The GeoTIFF format contains a set of embedded tags with metadata about the raster data. We can use the function GDALinfo() to get information about our raster data before we read that data into R. It is ideal to do this before importing your data.


```{r lookAtInto}
GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
```

We can store this information in R

```{r CaptureInfo}
HARV_dsmCrop_info <- capture.output(
  GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
)
```

Each line of text that was printed to the console is now stored as an element of the character vector HARV_dsmCrop_info. We will be exploring this data throughout this episode. By the end of this episode, you will be able to explain and understand the output above.

### Open a Raster in R

```{r OpenRaster}
DSM_HARV <- 
  raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

DSM_HARV
```


```{r Summary}
summary(DSM_HARV)
```

Note the warning - unless you force R to calculate these statistics using every cell in the raster, it will take a random sample of 100,000 cells and calculate from that instead. To force calculation on more, or even all values, you can use the parameter maxsamp:

```{r}
summary(DSM_HARV, maxsamp = ncell(DSM_HARV))
```

Convert it to a dataframe in order to plot it using ggplot2

```{r}
DSM_HARV_df <- as.data.frame(DSM_HARV, xy = TRUE)
```

Can see a standard dataframe format

```{r LookDF}
str(DSM_HARV_df)
```

Let's use ggplot to plot the raster. 

We will set the color scale to scale_fill_viridis_c which is a color-blindness friendly color scale. We will also use the coord_quickmap() function to use an approximate Mercator projection for our plots. This approximation is suitable for small areas that are not too close to the poles. Other coordinate systems are available in ggplot2 if needed, you can learn about them at their help page ?coord_map. (DEMO this here)

```{r DemoGG}
DSM_HARV_df %>%
  ggplot() +
  geom_raster(aes(x = x, y = y, fill = HARV_dsmCrop)) +   scale_fill_viridis_c() +
  coord_quickmap()
```

### View Raster Coordinate Reference System (CRS) in R:

We can view the CRS string associated with our R object using the crs() function.

```{r}
# Challenge: What units are our data in?
crs(DSM_HARV)
```


Understanding CRS in Proj4 Format:
proj=utm: the projection is UTM, UTM has several zones.
zone=18: the zone is 18
datum=WGS84: the datum is WGS84 (the datum refers to the 0,0 reference for the coordinate system used in the projection)
units=m: the units for the coordinates are in meters
ellps=WGS84: the ellipsoid (how the earth’s roundness is calculated) for the data is WGS84

Calculate Raster Min and Max Values
```{r}
minValue(DSM_HARV)
```
```{r}
maxValue(DSM_HARV)
```

Set min and max values:
```{r}
DSM_HARV <- setMinMax(DSM_HARV)
```

We can use the raster() function to import one single band from a single or multi-band raster. We can view the number of bands in a raster using the nlayers() function.
```{r}
nlayers(DSM_HARV)
```

To highlight NA values in ggplot, alter the scale_fill_*() layer to contain a colour instruction for NA values, like scale_fill_viridis_c(na.value = 'deeppink')

```{r}
GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
```

Bad Data Values in Rasters are different from NoDataValue(s)

Create A Histogram of Raster Values:
```{r}
ggplot() +
    geom_histogram(data = DSM_HARV_df, aes(HARV_dsmCrop))
```


```{r}
ggplot() +
    geom_histogram(data = DSM_HARV_df, aes(HARV_dsmCrop), bins = 40)
```



Notes on raster structure
