### Raster structure

Questions:
What is a raster dataset?
How do I work with and plot raster data in R?
How can I handle missing or bad data values for a raster?

Objectives:
Describe the fundamental attributes of a raster dataset.
Explore raster attributes and metadata using R.
Import rasters into R using the raster package.
Plot a raster file in R using the ggplot2 package.
Describe the difference between single- and multi-band rasters.


installing the raster and rgdal packages: install.packages("raster"). Note that for raster package you might need to install sp package using install.packages("sp")
as wel ass install.packages("rgdal")
```{r}
library(sp)
library(raster)
library(rgdal)
```

First download the data using the link below:
https://ndownloader.figshare.com/articles/2009586/versions/10

viewing the Raster file attributes
```{r}
GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
```

Storing this information in R
```{r}
HARV_dsmCrop_info <- capture.output(
  GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
)
```

Open a Raster in R
```{r}
DSM_HARV <- 
  raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

DSM_HARV
```

```{r}
summary(DSM_HARV)
```

```{r}
summary(DSM_HARV, maxsamp = ncell(DSM_HARV))
```

convert it to a dataframe in order to plot it using ggplot2
```{r}
DSM_HARV_df <- as.data.frame(DSM_HARV, xy = TRUE)
```

can see a standard dataframe format
```{r}
str(DSM_HARV_df)
```

Note: in case ggplot is not installed install.packages("ggplot2") or install.packages("ggplot2", dependencies = TRUE), then check the packages to make sure it is ticked
```{r}
ggplot() +
  geom_raster(data = DSM_HARV_df , aes(x = x, y = y, fill = HARV_dsmCrop)) +   scale_fill_viridis_c() +
  coord_quickmap()
```

View Raster Coordinate Reference System (CRS) in R:
We can view the CRS string associated with our R object using thecrs() function.
```{r}
crs(DSM_HARV)
```
Challenge: What units are our data in?

Understanding CRS in Proj4 Format:
proj=utm: the projection is UTM, UTM has several zones.
zone=18: the zone is 18
datum=WGS84: the datum is WGS84 (the datum refers to the 0,0 reference for the coordinate system used in the projection)
units=m: the units for the coordinates are in meters
ellps=WGS84: the ellipsoid (how the earth’s roundness is calculated) for the data is WGS84

Calculate Raster Min and Max Values
```{r}
minValue(DSM_HARV)
```
```{r}
maxValue(DSM_HARV)
```

Set min and max values:
```{r}
DSM_HARV <- setMinMax(DSM_HARV)
```

We can use the raster() function to import one single band from a single or multi-band raster. We can view the number of bands in a raster using the nlayers() function.
```{r}
nlayers(DSM_HARV)
```

To highlight NA values in ggplot, alter the scale_fill_*() layer to contain a colour instruction for NA values, like scale_fill_viridis_c(na.value = 'deeppink')

```{r}
GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
```

Bad Data Values in Rasters are different from NoDataValue(s)

Create A Histogram of Raster Values:
```{r}
ggplot() +
    geom_histogram(data = DSM_HARV_df, aes(HARV_dsmCrop))
```


```{r}
ggplot() +
    geom_histogram(data = DSM_HARV_df, aes(HARV_dsmCrop), bins = 40)
```



Notes on raster structure
