### Open and Plot Shapefiles in R

1. Questions:
  - How can I distinguish between and visualize point, line and polygon vector data?

2. Objectives

  - Know the difference between point, line, and polygon vector elements.
  - Load point, line, and polygon shapefiles into R.
  - Access the attributes of a spatial object in R.

The data we'll be using come from the NEON Harvard Forest field site, which we have been working with in previous episodes.

## Import shapefiles

- We will use the `sf` package to work with vector data in R. This stands for "Spatial features".

- `rgdal` automatically loads when `sf` is loaded. I'll mention here that there is another package called `sp`, which was the "old" one used to work with vector files in R. We will use `sf` in this course, as it is newer and more robust (as in, doesn't change the data type when doing things linke merges and joins) - but you sometimes will need to use `sp` with other R packages, most notably raster, - and you'll see it on stackoverflow. 


The shapefiles that we will import are:
- A polygon shapefile representing our field site boundary, which we'll call our Area Of Interest or AOI
- A line shapefile representing roads
- A point shapefile representing the location of the Fisher flux tower - which we will call "tower" from now on.


```{r loadshapefile}
library(sf)
# To import shapefiles we use the sf function st_read()
aoi_boundary_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HarClip_UTMZ18.shp")
```

Let's look at the metadata of our shapefile...


...

- Note that the spatial extent for a shapefile represents the combined extent for all spatial objects in the shapefile.


- There are commands to view this information from the R object as well

```{r}
st_geometry_type(aoi_boundary_HARV)
# you can see all possible geometry types using
levels(st_geometry_type(aoi_boundary_HARV))
# note some are more useful than others
# 
st_crs(aoi_boundary_HARV)

```

