## Explore and Plot by Shapefile Attributes

```{r reloadData, echo=F}
library(sf)
library(tidyverse)
theme_set(theme_minimal())
aoi_boundary_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HarClip_UTMZ18.shp")
lines_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARV_roads.shp")
point_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARVtower_UTM18N.shp")
```

We can use the "usual" R ncol and dim functions from R to look at our objects:

```{r}
dim(lines_HARV)
lines_HARV %>% head()
ncol(point_HARV)
```

```{r ch2}
# Explore the attributes associated with the point_HARV and aoi_boundary_HARV spatial objects.
# 
#     How many attributes does each have?
#     Who owns the site in the point_HARV data object?
# 
#     Which of the following is NOT an attribute of the point_HARV data object?
# 
# A) Latitude 
# B) County 
# C) Country
# 
```

Because this is R, we can use standard R operations to access specific columns of the sf dataframe:

```{r standardOps}
lines_HARV$TYPE
levels(lines_HARV$TYPE)

```
Furthermore, we can use dplyr commands to work with our object just like a dataframe - because that's what it is:

```{r}
class(lines_HARV)
```

Let's plot only the footpaths:

```{r}
lines_HARV %>%   ggplot() + geom_sf() 

lines_HARV %>% 
  filter(TYPE == "footpath") %>% 
  ggplot() + geom_sf() 
```

Now let's plot all of the different road types, colouring them in by type:

```{r}
# First, how many road types are there?
levels(lines_HARV$TYPE)


# let's build up the ggplot
lines_HARV %>%
  ggplot() + 
  geom_sf(aes(col = TYPE)) +
  scale_color_manual(values = c("red", "green", "blue", "magenta")) +
  ggtitle("NEON Harvard Forest Field Site", subtitle = "Roads Where Bikes and Horses Are Allowed") + 
  coord_sf()
```

```{r Ch3}
# Create a map of the state boundaries in the United States using the data located in your downloaded data folder: # NEON-DS-Site-Layout-Files/US-Boundary-Layers\US-State-Boundaries-Census-2014. 
# Apply a fill color to each state using its region value. Add a legend.

state_boundary_US <- 
st_read("data/NEON-DS-Site-Layout-Files/US-Boundary-Layers/US-State-Boundaries-Census-2014.shp")
levels(state_boundary_US$region)
colors <- c("purple", "springgreen", "yellow", "brown", "navy")
ggplot() +
  geom_sf(data = state_boundary_US, aes(color = region), size = 1) +
  scale_color_manual(values = colors) +
  ggtitle("Contiguous U.S. State Boundaries") + 
  coord_sf()
```

